Theoretical prediction of disrupted Min oscillation in flattened \emph{Escherichia coli}

### 2015-08-03 paper/plos.tex
add "changed" macro to highlight changes we made.

@@

* E-mail: Corresponding schuljef@physics.oregonstate.edu
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
  The dynamics of the Min-protein system help \emph{Escherichia coli}
  regulate the process of cell division by identifying the center of
  the cell.  While this system exhibits robust bipolar oscillations in
  wild-type cell shapes, recent experiments have shown that when the
  cells are mechanically deformed into wide, flattened out, irregular
  shapes, the spatial regularity of these oscillations breaks down. We
  employ widely used stochastic and deterministic models of the Min
  system to simulate cells with flattened shapes.  The deterministic
  model predicts strong bipolar oscillations, in contradiction with
  the experimentally observed behavior, while the stochastic model,
  which is based on the same reaction-diffusion equations,
  predicts more spatially irregular oscillations.
  We further report simulations of flattened but more symmetric
  shapes, which suggest that the flattening and lateral expansion may
  contribute as much to the irregular oscillation behavior as the
  asymmetry of the cell shapes.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author Summary}
  When an E. coli cell divides it must split into two relatively equal
  volumes in order to survive.  One mechanism that aids in this
  process is a system of proteins that oscillate between cell ends and
  help the original cell divide along its center.  A recent experiment
  found that by drastically deforming these cells this oscillation can
  be disrupted.  We examine these oscillations in similarly deformed
  cells computationally, using two widely used simulation models, and
  find good agreement using one of them.  The other model fails in a
  manner far more dramatic than has been previously observed.
  Furthermore, we find that the effect observed experimentally in
  asymmetric and flattened cells is primarily a result of the
  flattening and lateral expansion, not the asymmetry.

@@

\section*{Introduction}
It is vital that during the process of bacterial cell
division a cell avoid minicelling, or splitting into daughter cells
with lopsided volumes.  Instrumental to this process in
\emph{Escherichia coli} is a long FtsZ polymer chain that develops
near
the cell wall in the center region of the cell, helping dictate the
plane of division~\cite{adams2009bacterial,
  lutkenhaus2007assembly}. Previous experimental studies have shown
that the MinC protein, known to inhibit the FtsZ
polymer~\cite{shen2010examination}, exhibits regular pole to pole
oscillatory behavior between both ends of the wild-type pill-shaped
cell.  It thus has a higher time averaged concentration in the cell
poles than in the center region, which aides in preventing the FtsZ
from developing in the wrong region.  The MinC is recruited to these
poles by MinD, which itself interacts with another protein, MinE, in a
system which exhibits pole-to-pole oscillatory
behavior~\cite{shapiro2009and, yu1999ftsz,
  meacci2005min,huang2003dynamic,kerr2006division,mannik2009bacterial}.
%% \cite{hu1999topological, fu2001mine, shapiro2009and,
%%   yu1999ftsz, raskin1999rapid, meacci2005min, raskin1999minde}.

@@

A number of models of the MinD protein system have been developed that
accurately describe its basic oscillatory nature.
%
Early models involved free proteins that affect each others' rates of
diffusion and membrane attachment, but do not combine into compound
states~\cite{meinhardt2001pattern}.  Kruse extended this model by
incorporating recruitment and clustering of
MinD~\cite{kruse2002dynamic}.  In 2003 Huang improved upon Kruse's
approach with a simple and very successful simulation model that
incorporates MinD-MinE combination, ATPase hydrolysis, and MinD
membrane attachment that exhibits accurate MinD oscillations in
cylindrical cells~\cite{huang2003dynamic}. In this model cytoplasmic
MinD is recruited to the membrane by MinD that is already clustered
there (following observed non-linear attachment of MinD on the cell
membrane~\cite{hu2002dynamic,shih2002division}), and is stationary
once attached.
%

Several models \cite{bonny2013membrane,
  halatek2012highly} modify Huang's original model to more accurately
reflect experimental findings, which were published after his 2003
paper~\cite{meacci2006mobility, loose2011min}. These models introduce
diffusion along the membrane that is two to three orders of magnitude
slower than that within the cytoplasm, and increase the rate of
cytoplasmic diffusion by a factor of about five, reflecting an
experimental measurement of these
parameters~\cite{meacci2006mobility}.  They also increase
or assume instantaneous~\cite{bonny2013membrane} the nucleotide
exchange rate.  In addition, they revise the process of MinD
recruitment to the membrane, removing recruitment of MinD to the
membrane by MinD-MinE complexes, behavior that has not been observed
experimentally.  The model of Bonny \emph{et al.} further more allows
MinE to remain independently attached to the membrane after
disassociating with MinD~\cite{bonny2013membrane}.
%

There are only on the order of a thousand MinD proteins in a given
cell, which is fewer than the number of grid points used in our
simulations, suggesting that the deterministic continuum
description--which requires fractional numbers of proteins at each
grid point---may break down, and that stochastic behavior may play a
significant role.  To treat this, there have been a number of
applications of the Huang 2003 model that stochastically simulate the
same reaction-diffusion equations~\cite{fange2006noise,
  kerr2006division}.  These studies largely confirm the results of
Huang's deterministic model when applied to the wild-type, pill-shaped
phenotype.  However, the stochastic models are slightly more
successful in predicting experimentally observed oscillations in round
cell phenotypes~\cite{fange2006noise, huang2004min}, and they enable
prediction of fluctuations in the predicted
behavior~\cite{kruse2007experimentalist}, including the observation of
stochastic switching in short cells instead of a transition to static
behavior as would be predicted from a deterministic
model~\cite{fischer2010intra}.  Both deterministic and
stochastic models are widely used throughout systems
biology~\cite{lawson2013spatial, robb2014stochastic,
  oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
have unique advantages and limitations.  The deterministic approach
has an advantage in providing a simple prediction of average behavior,
while the stochastic approach enables prediction of fluctuations from
that mean, and reduces sensitivity to initial conditions.

%% biology~\cite{lawson2013spatial, robb2014stochastic,
%%   oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
In this paper we use Huang's original 2003
model~\cite{huang2003dynamic}, which has been widely
used~\cite{hattne2005stochastic,huang2004min,kerr2006division,
  fange2006noise} in both deterministic and stochastic variants, to
study flattened cells of $0.4\micron$ thickness that are similar to
those observed by M\"annik \emph{et al.}~\cite{mannik2009bacterial}.
This allows us to test the performance of this important model in more
extreme situations than has been done previously, furthering our
understanding of how and to what extent it breaks down.

\begin{figure}
  \includegraphics[width=8.7cm]{reactions}
  \caption{Reactions included in the Min system model of Huang
    \emph{et al.}, which is used in this
    work~\cite{huang2003dynamic}.}\label{fig:reactions}
\end{figure}

@@

Both Fig.~\ref{image-p} and its associated movies (see supplementary
material), show that the stochastic model exhibits a ``starry night''
effect characterized by spatially fixed points of protein density
build up, as recruitment leads to clusters of MinD forming on the wall
and then subsequently dissipating.  This result, which is not observed
in experiments, stems from the omission of diffusion along the
membrane from Huang's model, which would allow these recruitment
clusters to spread out.  Experiments have confirmed that diffusion
along the membrane does occur, albeit with a rate two orders of
magnitude lower than that in the cytoplasm~\cite{meacci2006mobility}.
The `stars' in the effect typically last for around 10 seconds. The
membrane-bound proteins will diffuse by a distance of around one
micron during this time, which is enough to spread out across the
polar section of the one micron diameter wild-type pill cells.  As we
see below, the flattened, irregular cells are large enough that this
one micron blurring might not completely eliminate this `starry night'
effect.

@@

\begin{figure*}
  \centering
  \hspace{-3cm}
  \vspace{1cm}
  \includegraphics[width=16cm]{../paper/plot-ave}
  \caption{Arrows depicting successive maxima in space
    and time overlayed on a color plot of the total MinD density
    averaged over the same time period.  The simulation time covered
    for the flattened cells is 350 seconds, which is the same
    period of time depicted in the experimental data plots of M\"annik
    \emph{et al.}~\cite{mannik2012robustness}.  For the wild-type pill
    shape, we only cover 250 seconds, in order to provide a
    useful comparison due to its shorter oscillation period.  The top
    row shows plots published by M\"annik \emph{et al.} of the MinD maxima behavior
    and the bottom two rows show our simulations using the stochastic
    and deterministic models, respectively.  We simulated
    approximations to the two shapes observed by M\"annik, which we call
    \emph{shape A} and \emph{shape B}.  In addition, we studied two
    flattened stadium shapes which we call \emph{stadium A} and
    \emph{stadium B} corresponding in aspect ratio and thickness to
    the two experimental shapes.  The spatial length scale of all the
    figures shown was identical. Each of the flattened cell shapes
    uses the same color scale for the number of proteins per unit
    area.  Finally, we display the natural pill shape, which was also
    featured in Figs.~\ref{image-p} and~\ref{corr-pill}, with a
    different color scale to reflect the thicker cell containing more
    proteins per cross-sectional area.  }
  \label{randst-plot-ave}
\end{figure*}

@@

In every case, including the wild-type pill shape, we see irregularity
in the location of the maxima when using the stochastic model.  The
deterministic model shows uniformly bipolar oscillation.  We conclude
that this model is \emph{not} chaotic (which would show irregularity
in the deterministic computation), but rather that the irregularity
results purely from stochastic processes.  From these results and how
they compare with experiment, we conclude that the deterministic model
is inadequate to explain experimental observations of the locations of
density maxima of the MinD protein.  We note, however, that it is
possible that there exists another deterministic model that does exhibit
chaotic behavior.

@@

The predictions of the stochastic model show some similarity to the
experimental results in terms of irregularity of the locations of
maxima, but the stochastic model displays significantly less
variation in the direction of oscillations than experiment, suggesting
inadequacies in the model.  We also note that the stadium shapes in
Fig.~\ref{randst-plot-ave} appear to have maxima location irregularity
that is qualitatively similar to our irregular shapes: in both
stochastic simulations the overall oscillation is largely bipolar, but
with irregularity in the locations of the maxima that form at each
end.

However Fig.~\ref{randst-plot-ave} leaves unclear the importance of
the flattening and accompanying lateral expansion of the cell versus
cell shape irregularity and asymmetry in the temporal regularity of
the oscillations.  Although in the movies the stadiums appear to
display, on average over long time scales, a somewhat more regular
oscillation than appears in the irregular shapes \emph{A} and
\emph{B}, the limited time range of the arrow plots (350 seconds) is
inadequate to make this comparison.  We therefore return to the
correlation function between the number of proteins in a segment at
the top of the cell and the number of proteins in a segment at the
bottom of the cell.  The ``top'' and ``bottom'' regions for computing
the correlation time were established by dividing the long axis into
thirds.  This resulted in regions at the top and bottom ranging from
19\% to 33\% of the cell volume for shapes \emph{A} and \emph{B}, with
the stadium shapes featuring 30\% of the volume in the top and bottom
regions.  Fig.~\ref{fig:corr-B} shows this plot for shape \emph{B} and
stadium \emph{B}.  Here we plot correlations of correlation times up
to 500 seconds, a biologically relevant timescale.  E. coli cell
doubling times vary widely depending on environmental conditions, but
will often be between 20 minutes to 100
minutes~\cite{pierucci1972chromosome}, while fluorescent microscopy
experiments have shown that the FtsZ ring builds itself with a half
life of roughly 30 seconds~\cite{stricker2002rapid}.

We see that both in the shape \emph{B} and stadium \emph{B}, the
deterministic model is perfectly periodic.  Both the deterministic and
stochastic model correlations have the same number of proteins in
their cells, and have been normalized with the same normalization
factor, so that they can be directly compared against each other.  The
stochastic simulation of shape \emph{B} has a coherence time of 5.3
periods, while the corresponding simulation of stadium \emph{B} has a
coherence time of 8.6 periods.
%
The correlation function for stadium \emph{A} is similar to that of
stadium~\emph{B}, with similar coherence time (13 periods, not shown),
while shape~\emph{A} also exhibits a similar coherence time (5.5
periods).  This confirms that the stadium shapes do indeed oscillate
somewhat more regularly than the irregular shapes, but both are
comparable to the wild-type pill shape.
%
From the stochastic correlation data, we conclude that irregularity of
shape leads to a decrease in temporal regularity of the oscillatory
behavior, although even the very distorted shape~\emph{B} exhibiting
only a factor of two greater decoherence.

\section*{Conclusion}
We find qualitative agreement between our stochastic simulations and
experiments showing disrupted bipolar MinD oscillation in asymmetric
flattened E. coli cells~\cite{mannik2012robustness}.  As observed
experimentally, our simulations predict that MinD maxima will form in
a spatially irregular sequence of locations, which follow an overall
bipolar trend.  This result builds upon existing results showing that
this model and its variants are effective in a variety of wild-type
and mutated cell shapes~\cite{fange2006noise, varma2008min,
  kruse2007experimentalist}, and reinforces that stochastic variations
of Huang's 2003 model~\cite{fange2006noise, kerr2006division} has
considerable predictive power beyond the standard wild-type cell.

In contrast to the stochastic model, the original deterministic
version of the model~\cite{huang2003dynamic} predicts behavior that
contradicts experimental observations.  Specifically, the
deterministic model predicts robust and periodic bipolar oscillation   
in irregularly shaped cells.  Thus we conclude that this method is
inconsistent with experiment, and cannot be relied upon for
predictions of the behavior of the MinD system.  This is the most
clear example of the inadequacy of the deterministic simulation method
in reproducing experimental behavior of the MinD system to date.
Until now the results of deterministic and stochastic simulation have
largely coincided, with the stochastic method showing minor
differences in behavior when compared to deterministic
models~\cite{kerr2006division, fange2006noise, huang2004min,
  kruse2007experimentalist}.  The regular oscillations in the
deterministic case also allow us to conclude that Huang's model is
non-chaotic.  Because of their similar construction, we anticipate
that recent variants of the model~\cite{bonny2013membrane,
  fange2006noise, halatek2012highly} may also prove to be
non-chaotic.

We find that flattened but regular and symmetric cells exhibit MinD
oscillations that are qualitatively similar to the oscillations
observed experimentally (and in our simulation) in irregular and
asymmetric cells.  This demonstrates that asymmetry is not required in
order to induce spatially irregular MinD oscillations.  In addition,
the temporal regularity of oscillations is only moderately affected by
irregularity in the shape of flattened cells.  Thus we conclude that
it may be the flattening of the cells, and the accompanying
expansion in the transverse direction, that causes the disruption of
regular bipolar oscillation which is observed by M\"annik \emph{et
  al.}~\cite{mannik2012robustness}.

>>>>

%\newcommand{\changed}[1]{{\color{red}#1}}
\newcommand{\changed}[1]{}

@@

* E-mail: Corresponding schuljef@physics.oregonstate.edu
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
  The dynamics of the Min-protein system help \emph{Escherichia coli}
  regulate the process of cell division by identifying the center of
  the cell.  While this system exhibits robust bipolar oscillations in
  wild-type cell shapes, recent experiments have shown that when the
  cells are mechanically deformed into wide, flattened out, irregular
  shapes, the spatial regularity of these oscillations breaks down. We
  employ widely used stochastic and deterministic models of the Min
  system to simulate cells with flattened shapes.  The deterministic
  model predicts strong bipolar oscillations, in contradiction with
  the experimentally observed behavior, while the stochastic model,
  which is based on the same reaction-diffusion equations,
  \changed{predicts more spatially irregular oscillations}.
  We further report simulations of flattened but more symmetric
  shapes, \changed{which suggest that the flattening and lateral expansion may
  contribute as much to the irregular oscillation behavior as the
  asymmetry of the cell shapes}.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author Summary}
  When an E. \changed{coli} cell divides it must split into two relatively equal
  volumes in order to survive.  One mechanism that aids in this
  process is a system of proteins that oscillate between cell ends and
  help the original cell divide along its center.  A recent experiment
  found that by drastically deforming these cells this oscillation can
  be disrupted.  We examine these oscillations in similarly deformed
  cells computationally, using two widely used simulation models, and
  find good agreement using one of them.  The other model fails in a
  manner far more dramatic than has been previously observed.
  Furthermore, we find that the effect observed experimentally in
  asymmetric and flattened cells is primarily a result of the
  flattening and lateral expansion, not the asymmetry.

@@

\section*{Introduction}
It is vital that during the process of bacterial cell
division a cell avoid minicelling, or splitting into daughter cells
with lopsided volumes.  Instrumental to this process in
\emph{Escherichia coli} is a long FtsZ polymer chain that develops
\changed{near}
the cell wall in the center region of the cell, helping dictate the
plane of division~\cite{adams2009bacterial,
  lutkenhaus2007assembly}. Previous experimental studies have shown
that the MinC protein, known to inhibit the FtsZ
polymer~\cite{shen2010examination}, exhibits regular pole to pole
oscillatory behavior between both ends of the wild-type pill-shaped
cell.  It thus has a higher time averaged concentration in the cell
poles than in the center region, which aides in preventing the FtsZ
from developing in the wrong region.  The MinC is recruited to these
poles by MinD, which itself interacts with another protein, MinE, in a
system which exhibits pole-to-pole oscillatory
behavior~\cite{shapiro2009and, yu1999ftsz,
  meacci2005min,huang2003dynamic,kerr2006division,mannik2009bacterial}.
%% \cite{hu1999topological, fu2001mine, shapiro2009and,
%%   yu1999ftsz, raskin1999rapid, meacci2005min, raskin1999minde}.

@@

A number of models of the MinD protein system have been developed that
accurately describe its basic oscillatory nature.
%
Early models involved free proteins that affect each others' rates of
diffusion and membrane attachment, but do not combine into compound
states~\cite{meinhardt2001pattern}.  \changed{Kruse extended this model by
incorporating recruitment and clustering of
MinD~\cite{kruse2002dynamic}.  In 2003 Huang improved upon Kruse's
approach} with a simple and very successful simulation model that
incorporates MinD-MinE combination, ATPase hydrolysis, and MinD
membrane attachment that exhibits accurate MinD oscillations in
cylindrical cells~\cite{huang2003dynamic}. In this model cytoplasmic
MinD is recruited to the membrane by MinD that is already clustered
there (following observed non-linear attachment of MinD on the cell
membrane~\cite{hu2002dynamic,shih2002division}), and is stationary
once attached.
%

Several models \changed{\cite{bonny2013membrane,
  halatek2012highly}} modify Huang's original model to more accurately
reflect experimental findings, which were published after his 2003
paper~\cite{meacci2006mobility, loose2011min}. These models introduce
diffusion along the membrane that is two to three orders of magnitude
slower than that within the cytoplasm, and increase the rate of
cytoplasmic diffusion by a factor of about five, reflecting an
experimental measurement of these
parameters~\cite{meacci2006mobility}.  They also increase
or assume instantaneous~\cite{bonny2013membrane} the nucleotide
exchange rate.  In addition, they revise the process of MinD
recruitment to the membrane, removing recruitment of MinD to the
membrane by MinD-MinE complexes, behavior that has not been observed
experimentally.  The model of Bonny \emph{et al.} further more allows
MinE to remain independently attached to the membrane after
disassociating with MinD~\cite{bonny2013membrane}.
%

There are only on the order of a thousand MinD proteins in a given
cell, which is fewer than the number of grid points used in our
simulations, suggesting that the deterministic continuum
description--which requires fractional numbers of proteins at each
grid point---may break down, and that stochastic behavior may play a
significant role.  To treat this, there have been a number of
\changed{applications} of the Huang 2003 model that stochastically simulate the
same reaction-diffusion equations~\cite{fange2006noise,
  kerr2006division}.  These studies largely confirm the results of
Huang's deterministic model when applied to the wild-type, pill-shaped
phenotype.  However, the stochastic models are slightly more
successful in predicting experimentally observed oscillations in round
cell phenotypes~\cite{fange2006noise, huang2004min}, and they enable
prediction of fluctuations in the predicted
behavior~\cite{kruse2007experimentalist}\changed{, including the observation of
stochastic switching in short cells instead of a transition to static
behavior as would be predicted from a deterministic
model~\cite{fischer2010intra}.}  Both deterministic and
stochastic models are widely used throughout systems
biology~\cite{lawson2013spatial, robb2014stochastic,
  oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
have unique advantages and limitations.  The deterministic approach
has an advantage in providing a simple prediction of average behavior,
while the stochastic approach enables prediction of fluctuations from
that mean, and reduces sensitivity to initial conditions.

%% biology~\cite{lawson2013spatial, robb2014stochastic,
%%   oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
In this paper we use Huang's original 2003
model~\cite{huang2003dynamic}, which has been widely
used~\changed{\cite{hattne2005stochastic,huang2004min,kerr2006division,
  fange2006noise}} in both deterministic and stochastic variants, to
study flattened cells of $0.4\micron$ thickness that are similar to
those observed by M\"annik \emph{et al.}~\cite{mannik2009bacterial}.
\changed{This allows us to test the performance of this important model in more
extreme situations than has been done previously, furthering our
understanding of how and to what extent it breaks down.}

\begin{figure}
  \includegraphics[width=8.7cm]{reactions}
  \caption{Reactions included in the \changed{Min system} model of Huang
    \emph{et al.}, which is used in this
    work~\cite{huang2003dynamic}.}\label{fig:reactions}
\end{figure}

@@

Both Fig.~\ref{image-p} and its associated movies (see supplementary
material), show that the stochastic model exhibits a ``starry night''
effect characterized by spatially fixed points of protein density
build up, as recruitment leads to clusters of MinD forming on the wall
and then subsequently dissipating.  \changed{This result, which is not observed
in experiments, stems from the omission of diffusion along the
membrane from Huang's model, which would allow these recruitment
clusters to spread out.}  Experiments have confirmed that diffusion
along the membrane does occur, albeit with a rate two orders of
magnitude lower than that in the cytoplasm~\cite{meacci2006mobility}.
The `stars' in the effect typically last for around 10 seconds. The
membrane-bound proteins will diffuse by a distance of around one
micron during this time, which is enough to spread out across the
polar section of the one micron diameter wild-type pill cells.  As we
see below, the flattened, irregular cells are large enough that this
one micron blurring \changed{might} not completely eliminate this `starry night'
effect.

@@

\begin{figure*}
  \centering
  \hspace{-3cm}
  \vspace{1cm}
  \includegraphics[width=16cm]{../paper/plot-ave}
  \caption{\changed{Arrows} depicting successive maxima in space
    and time overlayed on a color plot of the total MinD density
    averaged over the same time period.  The simulation time covered
    for the flattened cells is 350 seconds, which is the same
    period of time depicted in the experimental data plots of M\"annik
    \emph{et al.}~\cite{mannik2012robustness}.  For the wild-type pill
    shape, we only cover 250 seconds, in order to provide a
    useful comparison due to its shorter oscillation period.  The top
    row shows plots published by M\"annik \emph{et al.} of the MinD maxima behavior
    and the bottom two rows show our simulations using the stochastic
    and deterministic models, respectively.  We simulated
    approximations to the two shapes observed by M\"annik, which we call
    \emph{shape A} and \emph{shape B}.  In addition, we studied two
    flattened stadium shapes which we call \emph{stadium A} and
    \emph{stadium B} corresponding in aspect ratio and thickness to
    the two experimental shapes.  The spatial length scale of all the
    figures shown was identical. Each of the flattened cell shapes
    uses the same color scale for the number of proteins per unit
    area.  Finally, we display the natural pill shape, which was also
    featured in Figs.~\ref{image-p} and~\ref{corr-pill}, with a
    different color scale to reflect the thicker cell containing more
    proteins per cross-sectional area.  }
  \label{randst-plot-ave}
\end{figure*}

@@

In every case, including the wild-type pill shape, we see irregularity
in the location of the maxima when using the stochastic model.  The
deterministic model shows uniformly bipolar oscillation.  We conclude
that \changed{this} model is \emph{not} chaotic (which would show irregularity
in the deterministic computation), but rather that the irregularity
results purely from stochastic processes.  From these results and how
they compare with experiment, we conclude that the deterministic model
is inadequate to explain experimental observations of the locations of
density maxima of the MinD protein.  \changed{We note, however, that it is
possible that there exists another deterministic model that does exhibit
chaotic behavior.}

@@

The predictions of the stochastic model \changed{show some} similarity to the
experimental results in terms of irregularity of the locations of
maxima, \changed{but the stochastic model displays significantly less
variation in the direction of oscillations than experiment, suggesting
inadequacies in the model.}  We also note that the stadium shapes in
Fig.~\ref{randst-plot-ave} appear to have maxima location irregularity
that is qualitatively similar to our irregular shapes\changed{: in both
stochastic simulations the overall oscillation is largely bipolar, but
with irregularity in the locations of the maxima that form at each
end}.

However Fig.~\ref{randst-plot-ave} leaves unclear the importance of
the flattening and accompanying lateral expansion of the cell versus
cell shape irregularity and asymmetry in the temporal regularity of
the oscillations.  Although in the movies the stadiums appear to
display, on average over long time scales, a somewhat more regular
oscillation than appears in the irregular shapes \emph{A} and
\emph{B}, the limited time range of the arrow plots (350 seconds) is
inadequate to make this comparison.  We therefore return to the
correlation function between the number of proteins in a segment at
the top of the cell and the number of proteins in a segment at the
bottom of the cell.  \changed{The ``top'' and ``bottom'' regions for computing
the correlation time were established by dividing the long axis into
thirds.  This resulted in regions at the top and bottom ranging from
19\% to 33\% of the cell volume for shapes \emph{A} and \emph{B}, with
the stadium shapes featuring 30\% of the volume in the top and bottom
regions.}  Fig.~\ref{fig:corr-B} shows this plot for shape \emph{B} and
stadium \emph{B}.  Here we plot correlations of correlation times up
to 500 seconds, a biologically relevant timescale.  E. coli cell
doubling times vary widely depending on environmental conditions, but
will often be between 20 minutes to 100
minutes~\cite{pierucci1972chromosome}, while fluorescent microscopy
experiments have shown that the FtsZ ring builds itself with a half
life of roughly 30 seconds~\cite{stricker2002rapid}.

We see that both in the shape \emph{B} and stadium \emph{B}, the
deterministic model is perfectly periodic.  Both the deterministic and
stochastic model correlations have the same number of proteins in
their cells, and have been normalized with the same normalization
factor, so that they can be directly compared against each other.  The
stochastic simulation of shape \emph{B} has a coherence time of 5.3
periods, while the corresponding simulation of stadium \emph{B} has a
coherence time of 8.6 periods.
%
The correlation function for stadium \emph{A} is similar to that of
stadium~\emph{B}, with similar coherence time (13 periods, not shown),
while shape~\emph{A} also exhibits a similar coherence time (5.5
periods).  This confirms that the stadium shapes do indeed oscillate
somewhat more regularly than the irregular shapes, but both are
comparable to the wild-type pill shape.
%
From the stochastic correlation data, we conclude that irregularity of
shape \changed{leads to a decrease} in temporal regularity of the oscillatory
behavior, although even the very distorted shape~\emph{B} exhibiting
only a factor of two greater decoherence.

\section*{Conclusion}
We find \changed{qualitative} agreement between our stochastic simulations and
experiments showing disrupted bipolar MinD oscillation in asymmetric
flattened E. coli cells~\cite{mannik2012robustness}.  As observed
experimentally, our simulations predict that MinD maxima will form in
a spatially irregular sequence of locations, \changed{which follow an overall
bipolar trend}.  This result builds upon existing results showing that
this model and its variants are effective in a variety of wild-type
and mutated cell shapes~\cite{fange2006noise, varma2008min,
  kruse2007experimentalist}, and reinforces that stochastic variations
of Huang's 2003 model~\cite{fange2006noise, kerr2006division} has
considerable predictive power beyond the standard wild-type cell.

In contrast to the stochastic model, the original deterministic
version of the model~\cite{huang2003dynamic} predicts behavior that
contradicts experimental observations.  Specifically, the
deterministic model predicts robust and periodic bipolar oscillation   
in irregularly shaped cells.  Thus we conclude that this method is
inconsistent with experiment, and cannot be relied upon for
predictions of the behavior of the MinD system.  This is the most
clear example of the inadequacy of the deterministic simulation method
in reproducing experimental behavior of the MinD system to date.
Until now the results of deterministic and stochastic simulation have
largely coincided, with the stochastic method showing minor
differences in behavior when compared to deterministic
models~\cite{kerr2006division, fange2006noise, huang2004min,
  kruse2007experimentalist}.  The regular oscillations in the
deterministic case also allow us to conclude that \changed{Huang's} model is
non-chaotic.  Because of their similar construction, we anticipate
that recent variants of the model~\cite{bonny2013membrane,
  fange2006noise, halatek2012highly} \changed{may} also prove to be
non-chaotic.

We find that flattened but regular and symmetric cells exhibit MinD
oscillations that are qualitatively similar to the oscillations
observed experimentally (and in our simulation) in irregular and
asymmetric cells.  This demonstrates that asymmetry is not required in
order to induce spatially irregular MinD oscillations.  In addition,
the temporal regularity of oscillations is only moderately affected by
irregularity in the shape of flattened cells.  Thus we conclude that
it \changed{may be} the flattening of the cells, and the accompanying
expansion in the transverse direction, \changed{that causes} the disruption of
regular bipolar oscillation which is observed by M\"annik \emph{et
  al.}~\cite{mannik2012robustness}.
